New orbits of irregular satellites designed for the predictions of stellar occultations up to 2020, based on thousands of new observations

### 2016-07-08 Irregular-occ/mnras-source-revised/Irregular-occ.tex
.

\begin{abstract}
%Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system. It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin. The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects but this demands a very precise ephemeris based on new and good positions.
%\cite{GomesJunior2015} published 6523 observations obtained between 1992 and 2014 for 18 irregular satellites with an estimated error in the positions of about 60 to 80 mas.
%Based on this observations, we aimed to generate new numerical integration of the orbits of eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda. This is a short-time ephemeris that corrected systematic errors from previous ephemeris (JPL JUP300). For Phoebe we updated ephemeris from \cite{Desmars2013} using 75\% more positions than the previous one.
%Also Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018. This creates a good opportunity to predict stellar occultations due to the large number of events that can be predicted in these periods. Thus, we aimed to predict stellar occultations for the same 9 satellites cited above. Using the UCAC4 catalogue we managed to identify \noccs candidates of stellar occultations between the period of January, 2016 and December, 2020.
%We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.
\cite{GomesJunior2015} published \textbf{3613 positions for the 8 largest irregular satellites of Jupiter and 1787 positions for the largest irregular satellite of Saturn, Phoebe}.  These observations were made between \textbf{1995} and 2014 and have an estimated error of about 60 to 80 mas. Based on this set of positions, \textbf{we derived} new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda. %This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL jup300 for these bodies at least in the immediate future.
For Phoebe we updated the ephemeris from \cite{Desmars2013} using 75\% more positions than the previous one. Due to their orbital characteristics, it is common belief that the irregular satellites were captured by the giant planets in the early Solar System, but there is no consensus for a single model explaining where they were formed. Size, shape, albedo and composition would help to trace back their true origin, but these physical parameters are yet poorly known for irregular satellites. The observation of stellar occultations would allow for the determination of such parameters. Indeed Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018, improving a lot the chances of observing such events in the near future. Using the derived ephemerides and the UCAC4 catalogue we managed to identify \noccs candidate stellar occultations between January 2016 and December 2020 for the 9 satellites studied here. We discussed how the successful observation of a stellar occultation by these objects is possible and present some potential occultations.
\end{abstract}

@@

We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \citetalias{GomesJunior2015}%. We aim to eliminate the systematic errors in the JPL ephemeris
 (see Sec. \ref{Sec: integration}). %For an object at the distance of Jupiter, this represents an error langer than 700 km in the shadow path of an occultation.
For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff.

@@

\begin{table}
\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
\begin{center}
\begin{tabular}{lccc}
\hline  \hline
\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & mas$^ {a}$  & km & Ref. \tabularnewline
\hline
Ananke & 8 & 29 & 1 \tabularnewline
Carme & 13 & 46 & 1 \tabularnewline
Elara & 24 & 86 & 1 \tabularnewline
Himalia & 41 & $(150\times120) \pm 20^{b}$ & 2 \tabularnewline
Leda & 5 & 20 & 1 \tabularnewline
Lysithea & 10 & 36 & 1 \tabularnewline
Pasiphae & 17 & 62 & 1 \tabularnewline
Sinope & 10 & 37 & 1 \tabularnewline
\hdashline
Phoebe & 32 & $212 \pm 1.4^{b}$ & 3 \tabularnewline
%\hdashline
%Nereid & 15 & $340 \pm 50^{c}$ & 4 \tabularnewline
%Triton & 124 & $2707 \pm 2.0^{c}$ & 5 \tabularnewline
\hline
\end{tabular}
\end{center}
References: (1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}.\\
$^{a}${Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}\\
$^{b}${From Cassini observations.}
\par
\end{table}

%Since 2009 many successful observations of stellar occultations by TNOs have been reported in the literature \citep{Elliot2010, Sicardy2011, Ortiz2012, Braga-Ribas2013}, the main disadvantages in their prediction being large heliocentric distances and ephemeris error, facts somewhat compensated for the larger diameters involved. In contrast to TNOs, the irregular satellites have much better ephemeris because the orbits of their host planets are better  known, their observational time span is much wider and covers many orbital periods. Moreover, the irregular satellites are much closer to Earth which implies in a much smaller shadow path error in kilometers. These advantages may be somewhat balanced by the smaller sizes estimated for the irregular satellites.Thus, in comparison, the chances for a successful observation of an stellar occultation by an irregular satellite should be considered at least also as good as those by TNOs.

@@

\citetalias{GomesJunior2015} published \textbf{3613} precise positions for \textbf{the 8 largest irregular satellites of Jupiter} from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between \textbf{1995} and 2014.

@@

Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame. The satellites are submitted to the influence of the Sun and the \textbf{main bodies of the solar system (from Mercury to Pluto, plus the Moon)}, as well as those of the Galilean satellites and the first harmonics of Jupiter's gravity field. The axes of the reference frame are \textbf{expected to be those of the ICRS}.

We use the following notations: \begin{itemize}
\item \textbf{in one dynamical family consisting of $\mathcal{N}$ irregular satellites, $i$ will stand for the one whose equation of motion we are considering, $l$ will stand for another irregular satellite in gravitational interaction belonging to the same family}
\item $J$ Jupiter
\item \textbf{$j$ another body of the Solar System, among the Galilean satellites, the Sun, the planets, Pluto and the Moon (14 bodies)}
\item $M_j$ the mass of the $j$-th body, not an irregular satellite
\item $m_i$ the mass of the irregular satellite $i$
\item $\vec{r_i}$ the position of the $i$-th body with respect \textbf{to the center of Jupiter}
\item $r_{ij}$ the distance between bodies $i$ and $j$
\item $R_J$ the radius of Jupiter
\item $J_n$ the dynamic polar oblateness of the nth order for Jupiter's gravity field
\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
\item $\Phi_i$ is the \textbf{latitude} of the $i$-th satellite with respect to Jupiter's equator.
\end{itemize}

For an irregular satellite $i$, under the gravitational influence of Jupiter, the \textbf{seven other irregular satellites}, the regular Jovian satellites and\textbf{ the main bodies of the Solar System}, the equation of motion is:
\begin{equation}
\label{Eq:eq1}
\begin{array}{ll}

\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N} Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
&\displaystyle -\sum_{j=1}^{14}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
 & \displaystyle +GM_J \nabla U_{\bar{i}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
\end{array}
\end{equation}
where the last term in brackets and the last term in Eq. \ref{Eq:eq1} represent undirect perturbations. The oblateness potential seen by \textbf{a satellite $i$ because of Jupiter is (with a similar expression for the oblateness seen by a satellite $l$)}:
\begin{equation}\begin{array}{ll}

@@

\textbf{The expressions of $\nabla U$ have been developed in \cite{Lainey2004}}. \textbf{The equations of motion are integrated with the 15th order numerical integrator RADAU \citep{Everhart1985} using a constant step of one day}. \textbf{The positions of the objects of the solar system are provided by the DE423 ephemeris \citep{Folkner2010}, while those of the galilean satellites are provided by NOE2010 \citep{Lainey2004}}.
Our model was fitted to the observations through a least-squares procedure. The satellites were integrated one dynamical family at a time, to gain computing time, while losing minimum precision. Indeed, the interactions between satellites not belonging to the same dynamical family are negligible considering the short timespan of our integration.

The obtained ephemeris is hereafter referred to as STE, for Special-Tailored Ephemeris. The initial osculating elements at the origin of integration, \textbf{the number and time span of the observations of each satellite} are presented in Table \ref{Tab: sat_ell}.

@@

\textbf{Some methods to derive the errors of the ephemeris of irregular satellites can be found in \cite{Emelyanov2010}. In the Natural Satellites Ephemeride Server MULTI-SAT \citep{Emelyanov2008} the precisions can be obtained for the satellites at any given time from the \cite{Emelyanov2005} ephemeris updated to February 19, 2012. However, since the practical realization of the STE ephemeris is for help improving the prediction of stellar occultations by the irregular satellites in the immediate future, it is interesting to compare the STE with the other relevant ephemerides for the next few years.}

We compared the STE ephemeris to the JPL for all the Jupiter satellites we fitted, until 2021. For instance, the \textbf{maximum difference} between 2015 and 2021 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.

Fig. \ref{Fig: JPL-STE} displays the offsets of the positions published by \citetalias{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} jup300 JPL ephemeris and \cite{Emelyanov2005}\footnote{\textbf{Last update: February 19, 2012}} ephemeris. \textbf{The DE431 planetary ephemeris \citep{Folkner2014} was used to calculate the positions of Jupiter for the 3 models}. We see that the systematic JPL ephemeris offsets pointed out by \citetalias{GomesJunior2015} are reduced with our ephemeris, as expected.

\begin{figure}
\includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps}
\caption{Offsets in declination of the positions published by \protect\citetalias{GomesJunior2015} for Carme. The red "x" relate to the special-tailored ephemeris \textbf{(rms = 51)}, the blue "+" to the jup300 JPL ephemeris \textbf{(rms = 130)} and the green dot to \protect\cite{Emelyanov2005} \textbf{(rms = 92)}. As expected, the ephemeris \textbf{offsets} pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris. \label{Fig: JPL-STE}}
\end{figure}

In Table \ref{Tab:STE-error} we present the mean offsets and the respective standard deviation of the \citetalias{GomesJunior2015} positions relative to the same 3 ephemeris as above. We can see that the mean offsets as well as most of their standard deviations are greatly reduced with the STE ephemeris. \textbf{Of course, by construction, we should expect smaller offsets in the comparison of \citetalias{GomesJunior2015} positions with STE. However, it is not obvious that these offsets should be that smaller in comparison with the other ephemeris offsets. Notice that the \citetalias{GomesJunior2015} positions come from observations made with very distinct instruments at distant sites located at both Earth hemispheres (good parallax angle coverage), making this set of positions not remarkably distinct than any set of positions that were used in the construction of the other two ephemeris. Thus, these ephemeris offsets comparisons suggest that the accuracy of the STE ephemeris is at least slightly better than that of the other ephemeris, at least for the time span of our satellites' observations. This supports the utility of the STE ephemeris for the next few years, making it one of the best choices to use in stellar occultation predictions in the short future for these satellites.}

@@

%\textbf{The residuals obtained with STE are smaller than those obtained with JPL ephemeris, which was expected because the accuracy of an ephemeris decreases when we get further from the time of observations. If the ephemeris is fitted over too-short a timespan of observations, the initial conditions (positions, velocities, masses) used in the model will not be good enough and the model will differ greatly from the observed position of the satellite when we will accumulate new observations. The ephemeris is then said to be diverging.
%The main risk of divergence over time comes from the possible absence of long-term effects when fitting to a short timespan of observations. If that was the case, our ephemeris would diverge too quickly to be of any use.
%JPL ephemerides are fitted over all the available observations. As a result, they will diverge less quickly than our own. Though they are not as much precise as required for the use, they remain a precious reference to identify whether our own model presents a quick divergence.}

@@

For the specific case of Phoebe, the ninth satellite of Saturn, we have updated the ephemeris published in \cite{Desmars2013}. The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter. The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center \textbf{\citep{Arlot2009}}, and observations from Flagstaff \citep{NOFS} between 2012 and 2014. It represents a total number of 5886 observations from 1898 to 2014. In contrast, in \cite{Desmars2013} was used 3367 observations from 1898 to 2012. This represents an increase of almost 75\% in the number of observations, mainly with recent observations which is required for our purpose.

@@

We computed the precision of the PH15 ephemeris considering three sets of observations: all the positions available; only the positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.
\textbf{The precision is computed by propagation of the covariance matrix of the orbit determination process and by linear transformations giving the covariance matrix in spherical coordinates (right ascension and declination) at a specific date \citep[for more details, see][]{Desmars2013a}. This last matrix then provides the standard deviation in right ascension $\sigma_\alpha$ and declination  $\sigma_\delta$ at the required date, with $\sigma_s=\sqrt{\sigma^2_\alpha\cos^2\delta+\sigma^2_\delta}$ being the total error in the celestial sphere.}

In Fig. \ref{Fig:err-Phoebe} we show the comparison between them for the time span 2016-2021 in \textbf{on-sky Phoebe-Saturn angular separation}. It is possible to see that even considering only the positions of \citetalias{GomesJunior2015}, the estimated error of the ephemeris is smaller than 12 mas. The computed precision does not take into account the precision in the position of Saturn.

@@

%\subsection{Triton's and Nereid's ephemeris}
%
%Triton is an uncommon satellite. Its orbit is retrograde and inclined, but quasi-circular and very close to the planet compared to the irregular ones. Because Triton's orbit size is very small and its precession is not dominated by Solar perturbations, Triton is frequently excluded from the irregular satellites' class, but still studied together by many authors \citep{Sheppard2005, Jewitt2007}. Similarly to the irregular satellites, Triton was probably captured in the early solar system and may have the same origin as the TNOs \citep{Agnor2006}. However, Triton is bigger than the irregular satellites by an order of magnitude and has an atmosphere.
%
%For this satellite, we used the most recent ephemeris published by \cite{Emelyanov2015} which is based on more observations than previous ephemeris and covers 4 years more than the ephemeris of \cite{Jacobson2009}. In Fig. \ref{Fig:eph-Triton} we compare it with the ephemeris published in \cite{Zhang2014} and the NEP081 JPL ephemeris \citep{Jacobson2009}. The offsets between the three ephemeris are smaller than 20 mas for the period 2015-2021. This value is much smaller than the apparent size of Triton (124 mas, see Table \ref{Tab: satellite-diameter}), which indicates a good probability of observing an event by this object.
%
%\begin{figure*}
%\begin{centering}
%\subfigure[NEP081 JPL and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/JPL-EME_Triton.eps} \label{Fig:eph-Triton-eme}}
%\subfigure[\protect\cite{Zhang2014} and \protect\cite{Emelyanov2015}]{\includegraphics[width=8.7cm]{figures/Zhang-EME_Triton.eps} \label{Fig:eph-Triton-zhang}}
%\caption{Comparison between the \protect\cite{Emelyanov2015}, \protect\cite{Zhang2014} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Triton.}
%\label{Fig:eph-Triton}
%\end{centering}
%\end{figure*}
%
%For Nereid, we used the ephemeris published by \cite{Emelyanov2011}, which uses more recent observations than the JPL ephemeris published by \cite{Jacobson2009}. The comparison between the two ephemeris can be seen in Fig. \ref{Fig:eph-Nereid}. The right ascension offsets between them are smaller than 60 mas and the declination ones are smaller than 15 mas. For Nereid, ephemeris errors in right ascension translate to uncertainties in the central instant of a stellar occultation, so offsets of 60 mas are not critical. The declination ephemeris offsets are of the order of the apparent diameter (15 mas, see Table \ref{Tab: satellite-diameter}), resulting in fair chances (50\%) that the shadow crosses the expected Earth latitude predicted for an occultation. 
%
%\begin{figure}
%\begin{centering}
%\includegraphics[width=8.8cm]{figures/JPL-EME_Nereid.eps}
%\caption{Comparison between the \protect\cite{Emelyanov2011} and NEP081 JPL \protect\citep{Jacobson2009} ephemeris for the satellite Nereid.}
%\label{Fig:eph-Nereid}
%\end{centering}
%\end{figure}
%
%\cite{GomesJunior2015} published 902 new positions for Nereid from observations between 1992 and 2014. The development of a better ephemeris for Nereid using these new positions in the context discussed here of near future stellar occultations is desirable, but is out of the scope of this paper.

@@

A total of \noccs events were identified between January 2016 and December 2020. In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. It is possible to see an increase in the number of events found for Phoebe in 2018 and for the satellites of Jupiter in 2019-2020. This is because at that periods these satellites will cross the apparent galactic plane. We call attention that about 10\% of the events will involve stars brighter than \textbf{magnitude R=14}  (and almost 25\% brighter than R=15), which helps the attempt of amateur observers.

\begin{table}
\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2020.}
\begin{centering}
\begin{tabular}{lcccccc}
\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & 2016 & 2017 & 2018 & 2019 & 2020 & Total \tabularnewline
\hline
Ananke & 12 & 16 & 49 & 359 & 187 & 623 \tabularnewline
Carme & 20 & 14 & 30 & 369 & 220 & 653 \tabularnewline
Elara & 14 & 16 & 33 & 305 & 193 & 561 \tabularnewline
Himalia & 15 & 12 & 54 & 257 & 230 & 568 \tabularnewline
Leda & 8 & 24 & 38 & 362 & 208 & 640 \tabularnewline
Lysithea & 16 & 11 & 35 & 330 & 212 & 604 \tabularnewline
Pasiphae & 20 & 19 & 44 & 362 & 206 & 651 \tabularnewline
Sinope & 15 & 21 & 34 & 356 & 256 & 682 \tabularnewline
\hdashline
Phoebe & 32 & 98 & 238 & 79 & 13 & 460 \tabularnewline
%\hdashline
%Nereid & 11$^{a}$ & 1 & 1 & 0 & 0 & 13 \tabularnewline
%Triton & -- & 1 & 1 & 0 & 0 & 2 \tabularnewline
\hline
\end{tabular}
\par \end{centering}
\end{table}

Table \ref{Tab: sample-cds} shows a sample of the catalogue of predicted occultations and their parameters, which are necessary to produce occultation maps. Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS \textbf{(Centre de Données astronomiques de Strasbourg)}. In Fig. \ref{Fig: ocultacao} we show an example of an occultation map. This is an occultation by Elara that will happen in February 21, 2017. This event can be observed from Australia and it is one of the best opportunities for this object due to the slow velocity of the event and it involves a bright star (R*=12.4).

>>>>

\begin{abstract}
\cite{GomesJunior2015} published 3613 positions for the 8 largest irregular satellites of Jupiter and 1787 positions for the largest irregular satellite of Saturn, Phoebe.  These observations were made between 1995 and 2014 and have an estimated error of about 60 to 80 mas. Based on this set of positions, we derived new orbits for the eight largest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda. %This corrected systematic errors and resulted in an ephemeris more accurate/precise than JPL jup300 for these bodies at least in the immediate future.
For Phoebe we updated the ephemeris from \cite{Desmars2013} using 75\% more positions than the previous one. Due to their orbital characteristics, it is common belief that the irregular satellites were captured by the giant planets in the early Solar System, but there is no consensus for a single model explaining where they were formed. Size, shape, albedo and composition would help to trace back their true origin, but these physical parameters are yet poorly known for irregular satellites. The observation of stellar occultations would allow for the determination of such parameters. Indeed Jupiter will cross the galactic plane in 2019-2020 and Saturn in 2018, improving a lot the chances of observing such events in the near future. Using the derived ephemerides and the UCAC4 catalogue we managed to identify \noccs candidate stellar occultations between January 2016 and December 2020 for the 9 satellites studied here. We discussed how the successful observation of a stellar occultation by these objects is possible and present some potential occultations.
\end{abstract}

@@

We present in this paper new numerical integration of the orbits of the 8 major irregular satellites of Jupiter (Himalia, Elara, Pasiphae, Lysithea, Carme, Ananke, Sinope and Leda) using only the positions obtained by \citetalias{GomesJunior2015} (see Sec. \ref{Sec: integration}). For Phoebe, we updated the ephemeris of \cite{Desmars2013} using the observations of \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Center and observations from Flagstaff.

@@

\begin{table}
\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
\begin{center}
\begin{tabular}{lccc}
\hline  \hline
\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & mas$^ {a}$  & km & Ref. \tabularnewline
\hline
Ananke & 8 & 29 & 1 \tabularnewline
Carme & 13 & 46 & 1 \tabularnewline
Elara & 24 & 86 & 1 \tabularnewline
Himalia & 41 & $(150\times120) \pm 20^{b}$ & 2 \tabularnewline
Leda & 5 & 20 & 1 \tabularnewline
Lysithea & 10 & 36 & 1 \tabularnewline
Pasiphae & 17 & 62 & 1 \tabularnewline
Sinope & 10 & 37 & 1 \tabularnewline
\hdashline
Phoebe & 32 & $212 \pm 1.4^{b}$ & 3 \tabularnewline
\hline
\end{tabular}
\end{center}
References: (1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}.\\
$^{a}${Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}\\
$^{b}${From Cassini observations.}
\par
\end{table}

@@

\citetalias{GomesJunior2015} published 3613 precise positions for the 8 largest irregular satellites of Jupiter from observations made at the Observatório do Pico dos Dias (OPD), Observatoire Haute-Provence (OHP) and European Southern Observatory (ESO) between 1995 and 2014.

@@

Our numerical model describes the dynamical evolution of the irregular satellites of Jupiter in a jovicentric reference frame. The satellites are submitted to the influence of the Sun and the main bodies of the solar system (from Mercury to Pluto, plus the Moon), as well as those of the Galilean satellites and the first harmonics of Jupiter's gravity field. The axes of the reference frame are expected to be those of the ICRS.

We use the following notations: \begin{itemize}
\item in one dynamical family consisting of $\mathcal{N}$ irregular satellites, $i$ will stand for the one whose equation of motion we are considering, $l$ will stand for another irregular satellite in gravitational interaction belonging to the same family
\item $J$ Jupiter
\item $j$ another body of the Solar System, among the Galilean satellites, the Sun, the planets, Pluto and the Moon (14 bodies)
\item $M_j$ the mass of the $j$-th body, not an irregular satellite
\item $m_i$ the mass of the irregular satellite $i$
\item $\vec{r_i}$ the position of the $i$-th body with respect to the center of Jupiter
\item $r_{ij}$ the distance between bodies $i$ and $j$
\item $R_J$ the radius of Jupiter
\item $J_n$ the dynamic polar oblateness of the nth order for Jupiter's gravity field
\item $U_{\bar{l}\hat{J}}$ potential generated by the oblateness of Jupiter on the satellite $l$
\item $\Phi_i$ is the latitude of the $i$-th satellite with respect to Jupiter's equator.
\end{itemize}

For an irregular satellite $i$, under the gravitational influence of Jupiter, the seven other irregular satellites, the regular Jovian satellites and the main bodies of the Solar System, the equation of motion is:
\begin{equation}
\label{Eq:eq1}
\begin{array}{ll}

\ddot{\vec{r_i}}= & \displaystyle -GM_J\frac{\vec{r_J}-\vec{r_i}}{r_{iJ}^3}-\sum_{l=1,l\neq i}^\mathcal{N} Gm_l\frac{\vec{r_l}-\vec{r_i}}{r_{il}^3}\\
&\displaystyle -\sum_{j=1}^{14}GM_j \left(\frac{\vec{r_j}-\vec{r_i}}{r_{ij}^3} - \frac{\vec{r_j}-\vec{r_J}}{r_{Jj}^3} \right)\\
 & \displaystyle +GM_J \nabla U_{\bar{i}\hat{J}} -\sum_{l=1}^\mathcal{N} Gm_l\nabla U_{\bar{l}\hat{J}}
\end{array}
\end{equation}
where the last term in brackets and the last term in Eq. \ref{Eq:eq1} represent undirect perturbations. The oblateness potential seen by a satellite $i$ because of Jupiter is (with a similar expression for the oblateness seen by a satellite $l$):
\begin{equation}\begin{array}{ll}

@@

The expressions of $\nabla U$ have been developed in \cite{Lainey2004}. The equations of motion are integrated with the 15th order numerical integrator RADAU \citep{Everhart1985} using a constant step of one day. The positions of the objects of the solar system are provided by the DE423 ephemeris \citep{Folkner2010}, while those of the galilean satellites are provided by NOE2010 \citep{Lainey2004}.
Our model was fitted to the observations through a least-squares procedure. The satellites were integrated one dynamical family at a time, to gain computing time, while losing minimum precision. Indeed, the interactions between satellites not belonging to the same dynamical family are negligible considering the short timespan of our integration.

The obtained ephemeris is hereafter referred to as STE, for Special-Tailored Ephemeris. The initial osculating elements at the origin of integration, the number and time span of the observations of each satellite are presented in Table \ref{Tab: sat_ell}.

@@

Some methods to derive the errors of the ephemeris of irregular satellites can be found in \cite{Emelyanov2010}. In the Natural Satellites Ephemeride Server MULTI-SAT \citep{Emelyanov2008} the precisions can be obtained for the satellites at any given time from the \cite{Emelyanov2005} ephemeris updated to February 19, 2012. However, since the practical realization of the STE ephemeris is for help improving the prediction of stellar occultations by the irregular satellites in the immediate future, it is interesting to compare the STE with the other relevant ephemerides for the next few years.

We compared the STE ephemeris to the JPL for all the Jupiter satellites we fitted, until 2021. For instance, the maximum difference between 2015 and 2021 is at most 98 mas in $\Delta \alpha \cos \delta$ and 58 mas in $\Delta \delta$ for Himalia and 181 mas in $\Delta \alpha \cos \delta$ and 152 mas in $\Delta \delta$ for Carme.

Fig. \ref{Fig: JPL-STE} displays the offsets of the positions published by \citetalias{GomesJunior2015} for the satellite Carme in declination relative to the STE ephemeris, to \cite{Jacobson2012} jup300 JPL ephemeris and \cite{Emelyanov2005}\footnote{Last update: February 19, 2012} ephemeris. The DE431 planetary ephemeris \citep{Folkner2014} was used to calculate the positions of Jupiter for the 3 models. We see that the systematic JPL ephemeris offsets pointed out by \citetalias{GomesJunior2015} are reduced with our ephemeris, as expected.

\begin{figure}
\includegraphics[width=8.8cm]{figures/Carme_ephemeris.eps}
\caption{Offsets in declination of the positions published by \protect\citetalias{GomesJunior2015} for Carme. The red "x" relate to the special-tailored ephemeris (rms = 51), the blue "+" to the jup300 JPL ephemeris (rms = 130) and the green dot to \protect\cite{Emelyanov2005} (rms = 92). As expected, the ephemeris offsets pointed out by \protect\citetalias{GomesJunior2015} are reduced with the STE ephemeris. \label{Fig: JPL-STE}}
\end{figure}

In Table \ref{Tab:STE-error} we present the mean offsets and the respective standard deviation of the \citetalias{GomesJunior2015} positions relative to the same 3 ephemeris as above. We can see that the mean offsets as well as most of their standard deviations are greatly reduced with the STE ephemeris. Of course, by construction, we should expect smaller offsets in the comparison of \citetalias{GomesJunior2015} positions with STE. However, it is not obvious that these offsets should be that smaller in comparison with the other ephemeris offsets. Notice that the \citetalias{GomesJunior2015} positions come from observations made with very distinct instruments at distant sites located at both Earth hemispheres (good parallax angle coverage), making this set of positions not remarkably distinct than any set of positions that were used in the construction of the other two ephemeris. Thus, these ephemeris offsets comparisons suggest that the accuracy of the STE ephemeris is at least slightly better than that of the other ephemeris, at least for the time span of our satellites' observations. This supports the utility of the STE ephemeris for the next few years, making it one of the best choices to use in stellar occultation predictions in the short future for these satellites.

@@

@@

For the specific case of Phoebe, the ninth satellite of Saturn, we have updated the ephemeris published in \cite{Desmars2013}. The new ephemeris (PH15) used the same dynamical model, including the perturbations of the Sun and the eight planets, the eight major satellites of Saturn and the $J_2$ parameter. The observations used to fit the model are identical to \cite{Desmars2013} (including 223 Cassini observations) with additional observations from \citetalias{GomesJunior2015}, \cite{Peng2015}, observations from Minor Planet Circulars between 2012 and 2014 (available on the Natural Satellite Data Center \citep{Arlot2009}, and observations from Flagstaff \citep{NOFS} between 2012 and 2014. It represents a total number of 5886 observations from 1898 to 2014. In contrast, in \cite{Desmars2013} was used 3367 observations from 1898 to 2012. This represents an increase of almost 75\% in the number of observations, mainly with recent observations which is required for our purpose.

@@

We computed the precision of the PH15 ephemeris considering three sets of observations: all the positions available; only the positions of \citetalias{GomesJunior2015}; and all the positions without \citetalias{GomesJunior2015}.
The precision is computed by propagation of the covariance matrix of the orbit determination process and by linear transformations giving the covariance matrix in spherical coordinates (right ascension and declination) at a specific date \citep[for more details, see][]{Desmars2013a}. This last matrix then provides the standard deviation in right ascension $\sigma_\alpha$ and declination  $\sigma_\delta$ at the required date, with $\sigma_s=\sqrt{\sigma^2_\alpha\cos^2\delta+\sigma^2_\delta}$ being the total error in the celestial sphere.

In Fig. \ref{Fig:err-Phoebe} we show the comparison between them for the time span 2016-2021 in on-sky Phoebe-Saturn angular separation. It is possible to see that even considering only the positions of \citetalias{GomesJunior2015}, the estimated error of the ephemeris is smaller than 12 mas. The computed precision does not take into account the precision in the position of Saturn.

@@

@@

A total of \noccs events were identified between January 2016 and December 2020. In Table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. It is possible to see an increase in the number of events found for Phoebe in 2018 and for the satellites of Jupiter in 2019-2020. This is because at that periods these satellites will cross the apparent galactic plane. We call attention that about 10\% of the events will involve stars brighter than magnitude R=14  (and almost 25\% brighter than R=15), which helps the attempt of amateur observers.

\begin{table}
\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2020.}
\begin{centering}
\begin{tabular}{lcccccc}
\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & 2016 & 2017 & 2018 & 2019 & 2020 & Total \tabularnewline
\hline
Ananke & 12 & 16 & 49 & 359 & 187 & 623 \tabularnewline
Carme & 20 & 14 & 30 & 369 & 220 & 653 \tabularnewline
Elara & 14 & 16 & 33 & 305 & 193 & 561 \tabularnewline
Himalia & 15 & 12 & 54 & 257 & 230 & 568 \tabularnewline
Leda & 8 & 24 & 38 & 362 & 208 & 640 \tabularnewline
Lysithea & 16 & 11 & 35 & 330 & 212 & 604 \tabularnewline
Pasiphae & 20 & 19 & 44 & 362 & 206 & 651 \tabularnewline
Sinope & 15 & 21 & 34 & 356 & 256 & 682 \tabularnewline
\hdashline
Phoebe & 32 & 98 & 238 & 79 & 13 & 460 \tabularnewline
\hline
\end{tabular}
\par \end{centering}
\end{table}

Table \ref{Tab: sample-cds} shows a sample of the catalogue of predicted occultations and their parameters, which are necessary to produce occultation maps. Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS (Centre de Données astronomiques de Strasbourg). In Fig. \ref{Fig: ocultacao} we show an example of an occultation map. This is an occultation by Elara that will happen in February 21, 2017. This event can be observed from Australia and it is one of the best opportunities for this object due to the slow velocity of the event and it involves a bright star (R*=12.4).
